Unfolding the genome's 3D architecture:\\ \Large{Insights in {\em P. falciparum} genome organization}

### 2017-07-28 doc/manuscript/paper.tex
More progress on the article

\begin{abstract}
The development of new means to probe samples for the three-dimensional
structure of DNA paves the way for in depth and systematic analysis of genome
architecture, offering new insights in many biological processes. 3C-like
methods can now assess physical interactions between pairs of loci in a
genome-wide fashion, thus leading to the creation of contact maps. The
democratization of such protocols creates many new opportunities for
methodological development: how to infer 3D models from these contact maps?
Can such models help us gain insights in biological process?

Recent studies assessed {\em P. falciparum} (the deadliest of the 5
human malaria parasites) for its genome organization using 3C-like protocol.
With its small genomic size, fairly simple yet changing genomic organization
across its lifecyle and strong correlation between chromatin folding and gene
expression, this parasite is the ideal case study for applying and developing
methods to infer 3D models and use them for downstream analysis.

@@

In recent year, chromosome conformation captures-like method, broadly referred
to as Hi-C allowed to identify physical interactions between two regions in a
genome-wide fashion, yielding information on their relative spatial distance
in the nucleus. Hi-C has opened new avenues for more systematic analysis of
the three-dimensional folding of the genome, paving the way for a better
understanding of the relations between 3D structure and gene regulation,
replication timing, epigenetic changes as well as many other biological
processes. The 3D genome of a wide variety of organisms has been studied in
the past year, including many species of yeasts \citep{duan:three-dimensional,
burton:species-level, mizuguchi:cohesin-dependent}, bacteria
\citep{umbarger:three-dimensional}, flies \citep{sexton:three-dimensional},
plants \citep{feng:genome-wide, wang:genome-wide} and numerous human and mouse
cell lines \citep{lieberman-aiden:comprehensive, rao:3D}. Within the past few
years, two studies of the three-dimensional structure of {\em P. falciparum}
have been released.  \citet{lemieux:genome-wide} probed several strains of
{\em P. falciparum} to understand the link between 3D structure and the
complex regulation of \textit{var} genes, a family of genes involved the
invasion of red blood cell and also responsible for the parasite's great
capacity to evade our immune system. \citet{ay:three-dimensional} assayed the
3D structure of the genome of the parasite during three key stages of the
erythrocytic cycle. The relatively small size of the genome yet relatively
complicated genome architecture, the complex life cycle and the strong link
between chromatin folding, gene regulation and epigenetic places {\em P.
falciparum} as an interesting case study for developing new methods for better
understanding the properties of genome folding and its link to gene
regulation.

An important part of analyzing the three-dimensional structure of the genome
of {\em P. falciparum} (as well as many other organisms) relied on the
inferring accurate 3D models of the structure. These models were then used as
a preliminary step to various analysis, such as identifying colocalized
elements, distinguishing between open and closed chromatin and last but not
least, integrating with other sources of data, such as gene expression or
chromatin modification. In recent years, a profusion of methods were developed
for creating 3D models of the genome, either as standalone methods or with the
specific goal to better understand an organism, a biological process (such as
the inactivation of chromosome X) or the folding of a chromosome. These
methods broadly fall into two categories: ``model-based'' and ``data-driven''.
The former methods consider the polymer nature of DNA to leverage the
theoretical and computational work done in statistical physics of polymers to
build with as few assumptions as possible many chromosome conformations. Those
chromosome conformations are then used to compare against experimental data,
such as Hi-C contact count matrices, in order to iteratively improve the
models. These models offer mechanistical insights into the folding of DNA. The
latter approaches use the experimental data to infer 3D models, by typically
minimizing a cost function ensuring the models are as consistent as possible
with the data.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{schemas/schema.pdf}
\caption{The art of understanding 3D genome structure}
\label{Fig1}
\end{figure}

\section*{``Data-driven'' three-dimensional models of DNA from contact maps}

The development of genome-wide and high-throughput protocols to probe samples
for their 3D genome architecture naturally paved the way to systematic and in
depth study of the folding mechanisms of DNA. Over the past 10 years, the
challenge of building 3D models from contact maps have been tackled by a
plethora of methods, some developed for a particular organisms, others with
the hope to be generalizable to any data sets. Methods model chromosomes as a
series of beads, and attempt to place the beads in a 3D euclidean space to
well represent the contact map. ``data-driven'' methods broadly fall into two
categories: (i) \textit{consensus} approaches, that aim at inferring a unique
mean structures best representing the contact count data; (ii)
\textit{ensemble} methods that yield a population of structures.

@@

\citet{lesne:3d} proposes a different approach to convert counts into
distances: they borrow the concept of shortest path from graph theory to
create a distance matrix. A graph is constructed from the contact map by
considering each loci as a node. Loci seen interacting are connected through
an edge of weight $\frac{1}{c_{ij}}$. The authors then compute wish-distances
$\delta_{ij}$ as the shortest path between node $i$ and $j$ in the graph.
Constructing wish-distances in such a way has two advantages: (i) low contact
counts do not contribute much to the wish-distances; (ii) the resulting
wish-distances form a distance matrix, and thus an optimal solution to the MDS
can be found.

In addition of having subtle differences in the derivation of wish distances
(which can have important effects on the resulting structures),
the different methods also vary by the inclusion of weights to reflect
confidence in ``wish-distances'' \citep{ay:three-dimensional}, or constraints
to reflect prior knowledge on the structure. See table~\ref{table:mds_detail}
for a summary of the characteristic of each method.

@@

\fixme{maybe skip this section}

@@

{\em Ensemble approaches} aim at inferring a population of structures
representative of the contact count map. The methods fall into two distinct
categories: the first type of problem casts a restraint-based 
optimization problem and sample local minima of the function
\citep{bau:three-dimensional, umbarger:three-dimensional}, while the second
type proposes a statistical modeling of the problem and samples the posterior
distribution \citep{rousseau:three, hu:bayesian}. A parallel can be made
between the first group of methods and {\em Consensus} MDS type of approaches,
and the second with {\em Consensus} statistical based methods.

@@

\citet{umbarger:three-dimensional,bau:three-dimensional, kalhor:genome} models
chromosomes as a series of beads, linked by restraining oscillators. These
oscillators can be thought of as a ``force'' between beads so that they come
into contact or ensure a minimal or maximal distance between those. The models
includes three different types of restraints: (i) beads seen interacting are
restrained with harmonic oscillators of strengths derived from the contact
counts; (ii) adjacent beads are ensured to be neither too close nor too far
from one another; (iii) the last oscillators are activated only when
constraints are not fulfilled to \todo{XXX}.  This yields an optimization problem
with a large number of local minima, which the authors sample from by running
50,000 minimizations starting from random starts.

@@

Akin to {\em consensus} MDS-like method and {\em ensemble}-local minima
methods,  the first approach is to consider each contact as a constraint and
to formulate an under constrained optimization problem. A population of
structures satisfying the constraints can be found by sampling local
minimas.

\citet{paulsen:manifold} proposes a very different method, and formulate a
\textit{manifold based optimization}, where a low rank PSD matrix (and thus a
distance matrix) is optimized to be as close as possible to the sparse contact
count matrix. Applying classical MDS on this low rank PSD matrix then yields a
\textit{consensus} 3D model of the genome.

\fixme{hirata et al}

\begin{table}[ht!]
\begin{center}
\scriptsize
\begin{tabular}{rlccccc}
\hline
\multirow{2}{*}{\textbf{\footnotesize Publication}} & \multirow{2}{*}{\textbf{\footnotesize Name}} &
\textbf{\footnotesize Consensus or} 
&\multirow{2}{*}{\textbf{\footnotesize MDS-based}} & \textbf{\footnotesize Statistical} &
\multirow{2}{*}{\textbf{\footnotesize Available}} \\
& & \textbf{\footnotesize Ensemble}
 & & \textbf{\footnotesize model} & \\
\hline
\hline
\scriptsize{\cite{dekker:capturing}} & & C & \CHECK & & \\
\scriptsize{\cite{duan:three-dimensional}} &  & C & \CHECK & & \CHECK \\
\scriptsize{\cite{tanizawa:mapping}} & & C & \CHECK & & \\
\scriptsize{\cite{ay:three-dimensional}} & & C & \CHECK & & \\
\scriptsize{\cite{ben-elazar:spatial}} & & C & \CHECK & & \CHECK \\
\scriptsize{\cite{varoquaux:statistical}} & Pastis & C & & \CHECK & \CHECK\\
\scriptsize{\cite{bau:three-dimensional}} & & E & & &  \\
\scriptsize{\cite{umbarger:three-dimensional}} & & E  & & &\\
\scriptsize{\cite{zhang:inference}} & chromSDE & C &  \CHECK & & \CHECK\\
\scriptsize{\cite{peng:sequencing}} & autochrom3D & C &  \CHECK & & \CHECK\\
\scriptsize{\cite{rousseau:three}} & & E & & \CHECK & \CHECK\\
\scriptsize{\cite{hu:bayesian}} & Bach & E/C &  & \CHECK & \CHECK\\
\scriptsize{\cite{kalhor:genome}} & & E &   & &\\
\scriptsize{\cite{lesne:3d}} & ShRec3D & C  & \CHECK & & \CHECK \\
\scriptsize{\cite{trieu:large}} & & C & & & \\
\scriptsize{\cite{trieu:3D}} & & C & &  & \\
\scriptsize{\cite{trieu:MOGEN}} & MOGEN & C & & & \\
\scriptsize{\cite{nagano:single-cell}} & & E & & & \\
\scriptsize{\cite{paulsen:manifold}} & & C & \CHECK & & \CHECK \\
\hline
\end{tabular}
\end{center}
\caption{\bf A comparison of 3D inference methods}{\small In this table, we summarize
properties of published methods to infer the 3D structure of the genome: (1)
is it a consensus or a ensemble based inference? (2) Is it an MDS based
method? (3) or relies on a statistical modeling; (4) is the software
available or not (to the best of our knowledge).}
\end{table}

@@

\section*{The art of modeling genome architecture.}

``Data-driven'' methods, as presented in the last section, use the
experimental contact maps to infer models as consistent as possible with the
data. ``Model-driven'' tackle the 3D-modeling challenge exactly in the
opposite manner. Consider the following task. You have a tens of thousands of
randomly placed beads-on-strings. Can you find the smallest sets of
constraints such that these beads-on-a-string interact overall the same way as
a given contact map? This is the daunting task accomplished by
``model-driven'' approach: chromosomes are modeled as polymers (or random
self-excluding fibers) under a small number constraints such that a generated
contact maps from these models match as closely as possible the observed
contact maps. These ``model-driven'' approaches offer mechanistical insights
into the genome architecture.

>>>>

\begin{abstract}

The development of new means to probe samples for the three-dimensional
structure of DNA paves the way for in depth and systematic analysis of genome
architecture. 3C-like methods coupled with high-throughput sequencing can now
assess physical interactions between pairs of loci in a genome-wide fashion,
thus enabling the creation of genome-by-genome contact maps. The
democratization of such protocols creates many new opportunities for
methodological development: how to infer 3D models from these contact maps?
Can such models help us gain insights in biological process?

Several recent studies applied such protocol to {\em P. falciparum} (the
deadliest of the 5 human malaria parasites), assessing its genome organization
at different moment of its life cycle. With its small genomic size, fairly
simple yet changing genomic organization across its lifecyle and strong
correlation between chromatin folding and gene expression, this parasite is
the ideal case study for applying and developing methods to infer 3D models
and use them for downstream analysis.

@@

In recent year, chromosome conformation captures-like method, broadly referred
to as Hi-C allowed to identify physical interactions between two regions in a
genome-wide fashion, yielding information on their relative spatial distance
in the nucleus \citep{lieberman-aiden:comprehensive}. 
Hi-C has opened new avenues for more systematic analysis of
the three-dimensional folding of the genome, paving the way for a better
understanding of the relations between 3D structure and gene regulation,
replication timing, epigenetic changes as well as many other biological
processes \citep{de:dna, dixon:topological, rao:3d}. 
The 3D genome of a wide variety of organisms has been studied in
the past year, including many species of yeasts \citep{duan:three-dimensional,
burton:species-level, mizuguchi:cohesin-dependent}, bacteria
\citep{umbarger:three-dimensional}, flies \citep{sexton:three-dimensional},
plants \citep{feng:genome-wide, wang:genome-wide} and numerous human and mouse
cell lines \citep{lieberman-aiden:comprehensive, rao:3d, dixon:topological}. Within the past few
years, two studies of the three-dimensional structure of {\em P. falciparum}
have been released.  \citet{lemieux:genome-wide} probed several strains of
{\em P. falciparum} to understand the link between 3D structure and the
complex regulation of \textit{var} genes, a family of genes involved the
invasion of red blood cell and also responsible for the parasite's great
capacity to evade our immune system. \citet{ay:three-dimensional} assayed the
3D structure of the genome of the parasite during three key stages of the
erythrocytic cycle. The relatively small size of the genome yet relatively
complicated genome architecture, the complex life cycle and the strong link
between chromatin folding, gene regulation and epigenetic places {\em P.
falciparum} as an interesting case study for developing new methods for better
understanding the properties of genome folding and its link to gene
regulation \citep{ay:three-dimensional, lemieux:genome-wide, ay:multiple}.

An important part of analyzing the three-dimensional structure of the genome
of {\em P. falciparum} (as well as many other organisms) relied on the
inference of accurate 3D models of the structure
\citep{duan:three-dimensional, ay:three-dimensional, bau:three-dimensional}. These models were then used as
a preliminary step to various analysis, such as identifying colocalized
elements, distinguishing between open and closed chromatin and last but not
least, integrating with other sources of data, such as gene expression or
chromatin modification \citep{duan:three-dimensional, bau:three-dimensional,
ay:three-dimensional}. In recent years, a profusion of methods were developed
for creating 3D models of the genome, either as standalone methods or with the
specific goal to better understand an organism
\citep{umbarger:three-dimensional, ay:multiple, duan:three-dimensional} or
a biological process (such as
the inactivation of chromosome X) \citep{bau:three-dimensional, deng:bipartite}. These
methods broadly fall into two categories: ``model-based'' and ``data-driven''.
The former methods consider the polymer nature of DNA to leverage the
theoretical and computational work done in statistical physics of polymers to
build with as few assumptions as possible many chromosome conformations. Those
chromosome conformations are then used to compare against experimental data,
such as Hi-C contact count matrices, in order to iteratively improve the
models. These models offer mechanistical insights into the folding of DNA. The
latter approaches use the experimental data to infer 3D models, by typically
minimizing a cost function ensuring the models are as consistent as possible
with the data.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{schemas/schema.pdf}
\caption{Understanding 3D genome structure using contact maps}
\label{Fig1}
\end{figure}

\section*{Inferring three-dimensional models of DNA from contact maps}

The development of genome-wide and high-throughput protocols to probe samples
for their 3D genome architecture naturally paved the way to systematic and in
depth study of the folding mechanisms of DNA. Over the past 10 years, the
challenge of building 3D models from contact maps have been tackled by a
plethora of methods, some developed for a particular organisms, others with
the hope to be generalizable to any data sets \citep{dekker:capturing,
duan:three-dimensional, tanizawa:mapping, ay:three-dimensional,
peng:sequencing, lesne:3d, ben-elazar:spatial, zhang:inference, trieu:large,
trieu:MOGEN, trieu:3D, bau:three-dimensional, umbarger:three-dimensional,
rousseau:three, hu:bayesian}. Methods model chromosomes as a
series of beads, and attempt to place the beads in a 3D euclidean space to
well represent the contact map. ``data-driven'' methods broadly fall into two
categories: (i) \textit{consensus} approaches, that aim at inferring a unique
mean structures best representing the contact count data; (ii)
\textit{ensemble} methods that yield a population of structures.

@@

\citet{lesne:3d, hirata:three-dimensional} proposes a different approach to convert counts into
distances: they borrow concepts from graph theory to
create a distance matrix (either using shortest-path algorithm or Djikstra
algorithm). A graph is constructed from the contact map by
considering each loci as a node. Loci seen interacting are connected through
an edge of weight $\frac{1}{c_{ij}}$. The authors then compute wish-distances
$\delta_{ij}$ as the shortest path between node $i$ and $j$ in the graph.
Constructing wish-distances in such a way has two advantages: (i) low contact
counts do not contribute much to the wish-distances; (ii) the resulting
wish-distances form a distance matrix, and thus an optimal solution to the MDS
can be found.

In addition of having subtle differences in the derivation of wish distances
(which can have important effects on the resulting structures),
the different methods also vary by the inclusion of weights to reflect
confidence in ``wish-distances'' \citep{ay:three-dimensional}, or constraints
to reflect prior knowledge on the structure \citep{ay:three-dimensional,
dekker:capturing, duan:three-dimensional} (see Table~\ref{table:mds_detail}
for a summary of the characteristic of each method).

@@

@@

{\em Ensemble} approaches aim at inferring a population of structures
representative of the contact count map. The methods fall into two distinct
categories: the first type of problem casts a restraint-based 
optimization problem and sample local minima of the function
\citep{bau:three-dimensional, umbarger:three-dimensional}, while the second
type proposes a statistical modeling of the problem and samples the posterior
distribution \citep{rousseau:three, hu:bayesian}. In short, the former is the
{\em ensemble} version of MDS-based and {\em ad hoc} methods, while the former the
{\em ensemble} version of statistical based methods.

@@

\citet{umbarger:three-dimensional,bau:three-dimensional, kalhor:genome} models
chromosomes as a series of beads, linked by restraining oscillators. These
oscillators can be thought of as a ``force'' between beads so that they come
into contact or ensure a minimal or maximal distance between those. The models
includes two different types of restraints: (i) beads seen interacting are
restrained with harmonic oscillators of strengths derived from the contact
counts; (ii) adjacent beads are ensured to be neither too close nor too far
from one another.  This yields an optimization problem with a large number of
local minima, which the authors sample from by running 50,000 minimization
starting from random starts.

@@

Akin to {\em ensemble}-local minima methods,  the first approach is to
consider each contact as a constraint and to formulate an under constrained
optimization problem. A population of structures satisfying the constraints
can be found by sampling local minimas \citep{nagano:single-cell}.
Akin to {\em consensus method}, one can attempt to construct a distance
matrix, either through manifold-based optimization \citep{paulsen:manifold}
(by finding a low rank PSD approximation of the sparse contact map) or
akin to ShRec3D, through graph-theory considering the weighted graph of
interactions \citep{hirata:three-dimensional}. A classical MDS method applied
to such distance matrix then yields a \textit{consensus} 3D model of the genome.

\begin{table}[ht!]
\begin{center}
\scriptsize
\begin{tabular}{rlccccc}
\hline
\multirow{2}{*}{\textbf{\footnotesize Publication}} & \multirow{2}{*}{\textbf{\footnotesize Name}} &
\textbf{\footnotesize Consensus or} 
&\multirow{2}{*}{\textbf{\footnotesize MDS-based}} & \textbf{\footnotesize Statistical} &
\multirow{2}{*}{\textbf{\footnotesize Available}} \\
& & \textbf{\footnotesize Ensemble}
 & & \textbf{\footnotesize model} & \\
\hline
\hline
\scriptsize{\cite{dekker:capturing}} & & C & \CHECK & & \\
\scriptsize{\cite{duan:three-dimensional}} &  & C & \CHECK & & \CHECK \\
\scriptsize{\cite{tanizawa:mapping}} & & C & \CHECK & & \\
\scriptsize{\cite{ay:three-dimensional}} & & C & \CHECK & & \\
\scriptsize{\cite{ben-elazar:spatial}} & & C & \CHECK & & \CHECK \\
\scriptsize{\cite{varoquaux:statistical}} & Pastis & C & & \CHECK & \CHECK\\
\scriptsize{\cite{bau:three-dimensional}} & & E & & &  \\
\scriptsize{\cite{umbarger:three-dimensional}} & & E  & & &\\
\scriptsize{\cite{zhang:inference}} & chromSDE & C &  \CHECK & & \CHECK\\
\scriptsize{\cite{peng:sequencing}} & autochrom3D & C &  \CHECK & & \CHECK\\
\scriptsize{\cite{rousseau:three}} & & E & & \CHECK & \CHECK\\
\scriptsize{\cite{hu:bayesian}} & Bach & E/C &  & \CHECK & \CHECK\\
\scriptsize{\cite{kalhor:genome}} & & E &   & &\\
\scriptsize{\cite{lesne:3d}} & ShRec3D & C  & \CHECK & & \CHECK \\
\scriptsize{\cite{trieu:large}} & & C & & & \\
\scriptsize{\cite{trieu:3D}} & & C & &  & \\
\scriptsize{\cite{trieu:MOGEN}} & MOGEN & C & & & \\
\scriptsize{\cite{nagano:single-cell}} & & E & & & \\
\scriptsize{\cite{paulsen:manifold}} & MBO & C & \CHECK & & \CHECK \\
\scriptsize{\cite{hirata:three-dimensional}} & RPR & C & \CHECK & & \\
\hline
\end{tabular}
\end{center}
\caption{\bf A comparison of 3D inference methods}{\small In this table, we summarize
properties of published methods to infer the 3D structure of the genome: (1)
is it a consensus or a ensemble based inference? (2) Is it an MDS based
method? (3) or relies on a statistical modeling; (4) is the software
available or not (to the best of our knowledge).}
\end{table}

@@

\section*{The art of modeling genome architecture}

``Data-driven'' methods, as presented in the last section, use the
experimental contact maps to infer models as consistent as possible with the
data. ``Model-driven'' tackle the 3D-modeling challenge exactly in the
opposite manner. Consider the following task. You have a tens of thousands of
randomly placed beads-on-a-string. Can you find the smallest sets of
constraints such that these beads-on-a-string interact overall the same way as
a given contact map? This is the daunting task accomplished by
``model-driven'' approach: chromosomes are modeled as polymers (or random
self-excluding fibers) under a small number constraints such that a generated
contact maps from these models match as closely as possible the observed
contact maps. These ``model-driven'' approaches offer powerful mechanistical
insights into the genome architecture, but are difficult to build in practice:
each organism, cell type, and time point require hand crafted sets of
constrained, build by iteratively improving models.
